---
title: "data visualization"
author: "yanting"
date: "2023-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading library, echo = TRUE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

Loading data
```{r loading in data, echo=TRUE}
data <- read_csv("../data/example_data_for_visualization.csv")
head(data)
unique(data$pos)
```

Trying to plot the vot distribution of different consonants according to their place of articulation:

```{r boxplot, echo=TRUE}
boxplot_data <- select(data,
                    subj,
                   gender,
                   vot, 
                   poa)

ggplot(boxplot_data, aes(x = poa, y = vot, fill = poa)) +
  geom_boxplot() +
  scale_fill_manual(values = c("orange", "lightblue", "pink")) +
  labs(y = "vot", x = "poa")

```

Trying to plot the histogram of words with different number of syllables for each consonant:

```{r histogram, echo=TRUE}
histogram_data <- select(data,
                   stop,
                   syll)

histogram_data_count <- histogram_data %>% 
  group_by(stop, syll) %>%
  summarize(count = n())

ggplot(histogram_data_count, aes(x = stop, fill = syll)) + 
  geom_bar(position = "dodge") +
  labs(x = "stop", y = "count of words", fill = "number of syllables") 


```

Trying to create a scatter plot for the usef0 for different vowels produced by speakers of different genders:

```{r scatterplot, echo=TRUE}
scatterplot_data <- select(data,
                          subj,
                          gender,
                          vowel,
                          usef0)

scatterplot_data_grouped <- group_by(
  scatterplot_data, gender, vowel
)
scatterplot_data <- summarize(scatterplot_data_grouped, f0 = mean(usef0, na.rm = TRUE))

ggplot(scatterplot_data, aes(x = vowel, y = f0, color = vowel)) + 
  geom_point(pos = position_dodge(width = 0.5), size = 1) +
  facet_wrap(~gender) +
  labs(x = "vowel", y = "f0")
```

I feel that this visualization is not very informative though... looking forward to learning more about how to effectively present data!