---
title: "data visualization"
author: "yanting"
date: "2023-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loading library, echo = TRUE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

Loading data
```{r loading in data, echo=TRUE}
data <- read_csv("../data/example_data_for_visualization.csv")
head(data)
unique(data$pos)
```

Trying to plot the vot distribution of different consonants according to their place of articulation:

```{r boxplot, echo=TRUE}
boxplot_data <- select(data,
                    subj,
                   gender,
                   vot, 
                   poa)
boxplot_data$poa <- factor(boxplot_data$poa , levels = c("lab", "cor", "dor"))

ggplot(boxplot_data, aes(x = poa, y = vot, fill = poa)) +
  geom_boxplot() +
  scale_fill_manual(values = c("orange", "lightblue", "pink")) +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c("labial", "coronal", "dorsal")) + 
  labs(y = "vot", x = "place of articulation", caption = "Vot of different stops plotted according to the place of articulation. The first quartile, the median and the third quartile are shown.") + 
  theme(plot.caption = element_text(hjust = 0.5))

```

Trying to plot the histogram of words with different number of syllables for each consonant (and I just realized that this is probably controlled for after seeing the plot...):

```{r histogram, echo=TRUE}
histogram_data <- select(data,
                   stop,
                   syll)
histogram_data$syll <- factor(histogram_data$syll, levels = c("one", "two", "more"))

histogram_data_count <- histogram_data %>% 
  group_by(stop, syll) %>%
  summarize(count = n())

ggplot(histogram_data_count, aes(x = stop, fill = syll)) + 
  geom_bar(position = "dodge") +
  scale_y_continuous(breaks = 0:2) + 
  scale_fill_manual(values = c("lightblue", "royalblue", "darkblue")) +
  labs(x = "stop", y = "count of words", fill = "number of syllables", caption = "Count of words with one, two or more syllables, grouped by the stop.") + 
  theme(plot.caption = element_text(hjust = 0.5))

```

Trying to create a scatter plot for the usef0 for different vowels produced by speakers of different genders:

```{r scatterplot, fig.width=10, fig.height = 4, echo=TRUE}
scatterplot_data <- select(data,
                          subj,
                          gender,
                          vowel,
                          usef0)

scatterplot_data_grouped <- group_by(
  scatterplot_data, gender, vowel
)
scatterplot_data <- summarize(scatterplot_data_grouped, f0 = mean(usef0, na.rm = TRUE))

ggplot(scatterplot_data, aes(x = vowel, y = f0, color = vowel)) + 
  geom_point(pos = position_dodge(width = 0.5), size = 1) +
  theme(legend.position = "none") +
  facet_wrap(~gender, labeller = labeller(facet_var = c("F" = "Female", "M" = "Male"))) +
  labs(x = "vowel", y = "f0", caption = "Average f0 of each vowel produced by female vs. male speakers.") + 
  theme(plot.caption = element_text(hjust = 0.5)) 

```
Ideally it would be nice to rename the variables on the x axis with the IPA of each vowel but I am not sure about what exact vowels they represent, so they are left as is for now.

Trying to create a bar chart for the average vowel duration for different vowels:

```{r barchart, echo=TRUE}
bar_data <- select(data,
                   vowel,
                   vdur)
bar_data_summary <- bar_data %>%
  group_by(vowel) %>%
  summarize(avg_duration = mean(vdur),
            sd_duration = sd(vdur),
            n = n()) 

ggplot(bar_data_summary, aes(x = reorder(vowel, avg_duration), y = avg_duration, fill = vowel)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = avg_duration - sd_duration,
                    ymax = avg_duration + sd_duration),
                width = 0.4, color = "black", size = 0.7) +
  theme(legend.position = "none") +
  labs(x = "Vowel", y = "Duration", caption = "Vowel duration from short to long for each vowel with error bar.") + 
  theme(plot.caption = element_text(hjust = 0.5))

```